On the Cram\'{e}r-Rao bound for polynomial phase signals

### 2012-10-17 paper.tex
added to paper and defs

\begin{abstract}
Polynomial-phase signals have numerous applications including radar, sonar, geophysics, and radio communication.  Of particular importance is the estimation of the parameters of a polynomial phase signal from a sequence of noisy observations.  Assuming that the noise is additive and Gaussian,  Peleg and Porat approximated the Cram\'{e}r-Rao bound for this estimation problem.  The approximation is accurate when the number of observations is sufficiently larger than the order of the polynomial phase signal.  In this paper we derive closed form formula for the exact Cram\'{e}r-Rao bounds.  This is achieved by making use of a family of orthogonal polynomials.
\end{abstract}

@@

A uniformly sampled, constant amplitude, polynomial-phase signal of order $m$ has the model~\cite{Peleg_DPT_1995, Peleg1991_CRB_PPS_1991, Peleg1991_est_class_PPS_1991, Rihaczek1996},
\[
s_n(b) = \alpha \exp\left( j \sum_{k = 0}^{m}{b_k (\Delta n)^k}\right)
\]
where $n \in \ints$, $\alpha$ is the signal amplitude (a positive real number), $b=(b_0, \dots, b_m)$ is a vector containing the polynomial coefficients (each a real number),  and $\Delta$ is the interval between consecutive samples.  We will assume, without loss of generality, that $\Delta = 1$.

Polynomial-phase signals have numerous applications including radar, sonar, geophysics, and radio communication~\cite{Angeby_estimating_2000}.  The signals are also used to describe the sounds emitted by bats for echo-location \cite{Peleg_DPT_1995}.  Of practical importance is the estimation of the coefficients $\beta = (\beta_0, \dots \beta_m)$ from $N$ consecutive samples $y_1, \dots. y_N$ where 
\[
y_n = s_n(\beta) + w_n
\]
and $\{w_n, n \in \ints\}$ is a sequence of noise variables.  Many estimators have been studied and implemented~\cite{Peleg_DPT_1995, Angeby_estimating_2000, Djuric_phase_unwrap_chirp_1990, Oshea_iterative_1996, Barbarossa_analysis_of_PPS_1997, Slocumb_polynomial_1994, Morelande_bayes_unwrapping_2008, Kitchen_polyphase_unwrapping_1994}.

Under the assumption that $\{w_n\}$ are additive white and Gaussian, Peleg and Porat~\cite{Peleg1991_CRB_PPS_1991} have derived the Cram\'{e}r-Rao lower bound for unbiased estimators of the coefficients $\beta$.  Their approach requires the inversion of a poorly conditioned Fisher information matrix.  To avoid this, they provide an approximation that is valid when the number of observations $N$ is much larger than the order of the polynomial phase signal $m$.  Here we remove the need for this approximation.  By converting the polynomial basis to one that is orthogonal we derive closed form formula for the inverse Fisher information matrix.  This leads to closed form formula for the Cram\'{e}r-Rao bounds.  %At the same time we tie together some existing literature on the Cram\'{e}r-Roa bound for polynomial phase signals~\cite{Peleg1991_CRB_PPS_1991,Ristic_crbpelegcomment_1998}.

The paper is organised as follows.  In Section~\ref{sec_crb} we derive the Fisher information matrix in a similar way to Peleg and Porat~\cite{Peleg1991_CRB_PPS_1991}.  The matrix is poorly conditioned and inverting it to obtain the Cram\'{e}r-Rao bound is difficult.  In Section~\ref{sec:changing-basis} we describe some general results regarding changing the basis in which a polynomial phase signal may be represented.  We describe the Fisher information matrix corresponding with a chosen basis, and we relate the matrices between two bases.  In Section~\ref{sec:discr-orth-polyn} we introduce the family of discrete orthogonal polynomials.  By transforming into this basis, the corresponding Fisher information matrix becomes diagonal, making matrix inversion simple.  This leads directly to closed form expressions for the Cram\'{e}r-Rao bounds.  To keep our derivation simple we initially assume that the amplitude $\alpha$ is a fixed constant, and not a parameter to be estimated.  However, including the amplitude is straightforward, and we do this in Section~\ref{sec:including-amplitude}.  Section~\ref{sec:simuations} presents the results of Monte-Carlo simulations that agree with our exact bounds.

@@

We make the assumption that the noise sequence $\{ w_n \}$ is independent and identically distributed complex Gaussian, each element having independent real and imaginary parts with zero mean and variance $\sigma^2$.  The probability density function of $w_n$ is 
\[
p(x) = \frac{1}{2 \pi \sigma^2} \exp\left( -\frac{\abs{x}^2}{2\sigma^2} \right)
\]
where $\abs{x}^2$ is the squared magnitude of the complex number $x$.  We wish to estimate $\beta$ from $y_1, \dots, y_N$.  The likelihood function is
\[
L(b) = \frac{1}{(2 \pi \sigma^2)^N} \exp\left( - \frac{1}{2\sigma^{2}} \sum_{n=1}^N \abs{y_n -  s_n(b) }^2 \right).
\]
%The maximum likelihood estimators of $\alpha, \beta_0, \dots, \beta_m$ are the maximisers of $L$.
% , that is
% \[
% (\hat{\alpha}, \hat{\beta}_0, \dots, \hat{\beta}_m) = \arg\max_{a \in \reals, (b_0, \dots, b_m) \in B} L(a, b_0, \dots, b_m) 
% \]  
% where $B \subset \reals^{m+1}$ describes a region of polynomial phase coefficients for which the estimation problem is identifiable~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase,McKilliam_LSU_polyest_part1_2012}.  We assume that $(\beta_0, \dots, \beta_m) \in B$.
For $i,k \in \{0, \dots, m\}$ define functions 
\[
f_{ik}(b) = f_{ki}(b) = -\expect \frac{ \partial^2 \log L(b) }{ \partial b_{i} \partial b_{k}}.
\]  
The Fisher information matrix $F$ is the $m+1$ by $m+1$ matrix with elements $F_{i k} = f_{ik}(\beta)$.  The elements are
\[
F_{ik} = F_{ki} = \frac{\alpha^2}{\sigma^2}\sum_{n=1}^{N} n^{i}n^{k}.
\]

Let $\hat{\beta}$ be an unbiased estimator of $\beta$.  The Cram\'{e}r-Rao bound asserts:  If $C$ is the covariance matrix of $\hat{\beta}$, then $C - F^{-1}$ is positive semidefinite, where $F^{-1}$ is the inverse of $F$.  The diagonal elements of a positive semidefinite matrix are nonnegative so $C_{ii} \geq F^{-1}_{ii}$ and correspondingly, 
\[
C_{ii} = \var \hat{\beta}_{i} \geq F_{ii}^{-1}, \qquad i=0,\dots,m.
\]  
Be aware that $F_{ik}^{-1}$ is the $ik$th element of the inverse matrix $F^{-1}$.  It is not the reciprocal of $F_{ik}$.  Peleg and Porat~\cite{Peleg1991_CRB_PPS_1991} found that $F$ is poorly conditioned and for this reason $F^{-1}$ is difficult to compute.  They give an approximation that is accurate when $N$ is sufficiently larger than $m$.  We will circumvent this problem by applying a change of basis to the polynomial coefficients $\beta$.  This will lead to closed form formula for $F^{-1}$.

@@

The likelihood function for $\gamma$ is 
\[
L_\gamma(c) = L(cP) = L(b)
\] 
and we can compute the corresponding Fisher information matrix.  It is found to be the $m+1$ by $m+1$ matrix $H$ with elements
\[
H_{ik} = H_{ki} = \frac{\alpha^2}{\sigma^2}\sum_{n=1}^{N} p_{i}(n) p_{k}(n).
\]
The Cram\'{e}r-Rao bound for $\gamma$ is found by taking the inverse of $H$.  The next lemma relates the two Fisher information matrices $H$ and $F$.

\begin{lemma}
Suppose we desire to estimate $\beta \in \reals^m$ and that $L(b)$ is the corresponding likelihood function and $F$ the corresponding $m \times m$ Fisher information matrix.  Let $P$ be an $m \times m$ invertible matrix and $\gamma \in \reals^m$ be such that $\beta = \gamma P$.  Then the likelihood function corresponding to $\gamma$ is $L_\gamma(c) = L(cP)$ and the corresponding Fisher information matrix is $H = P F P^\prime$. 
\end{lemma}
\begin{IEEEproof}
Put $b = cP$.  That $L_\gamma(c) = L(cP) = L(b)$ is plain.  The elements $F_{ik} = f_{ik}(\beta)$ where
\[
f_{ik}(b) = -\expect \frac{\partial \log L(b)}{\partial b_i \partial b_k}.
\]
The elements $H_{ik} = h_{ik}(\gamma)$ where
\[
h_{ik}(c) = -\expect \frac{\partial \log L_\gamma(c)}{\partial c_i \partial c_k} =  -\expect \frac{\partial \log L(cP)}{\partial c_i \partial c_k}.
\]
Since $\frac{\partial b_s}{\partial c_i} = P_{is}$ applying the chain rule yields
\begin{align*}
h_{ik}(c) &= -\expect \sum_{s=0}^{m-1} P_{is}\frac{\partial \log L(cP)}{\partial b_s \partial c_k} \\
&= -\expect \sum_{s=0}^{m-1}\sum_{t=0}^{m-1} P_{is} P_{kt}   \frac{\partial \log L(b)}{\partial b_s \partial b_t} \\
&= \sum_{s=0}^{m-1}\sum_{t=0}^{m-1} P_{is} P_{kt}  f_{st}(b).
\end{align*}
Putting $c = \gamma$ so that $b = \gamma P = \beta$ we have
\[
H_{ik} = \sum_{s=0}^{m-1}\sum_{t=0}^{m-1} P_{is} P_{kt}  F_{st},
\]
or equivalently $H = P F P^\prime$.
\end{IEEEproof}

>>>>

\begin{abstract}
Polynomial-phase signals have applications including radar, sonar, geophysics, and radio communication.  Of particular importance is the estimation of the parameters of a polynomial phase signal from a sequence of noisy observations.  Assuming that the noise is additive and Gaussian,  Peleg and Porat approximated the Cram\'{e}r-Rao bound for this estimation problem.  The approximation is accurate when the number of observations is sufficiently larger than the order of the polynomial phase signal.  In this paper we derive closed form formula for the exact Cram\'{e}r-Rao bounds.  This is achieved by making use of a family of orthogonal polynomials.
\end{abstract}

@@

A uniformly sampled, constant amplitude, polynomial-phase signal of order $m$ has the model~\cite{Peleg_DPT_1995, Peleg1991_CRB_PPS_1991, Peleg1991_est_class_PPS_1991, Rihaczek1996},
\[
s_n(b) = \alpha \exp\left( j \sum_{k = 0}^{m}{b_k (\Delta n)^k}\right)
\]
where $n \in \ints$, $\alpha$ is the signal amplitude (a positive real number), the argument $b=(b_0, \dots, b_m)$ is a vector representing the polynomial coefficients (each a real number),  and $\Delta$ is the interval between consecutive samples.  We will assume, without loss of generality, that $\Delta = 1$.

Polynomial-phase signals have applications including radar, sonar, geophysics, and radio communication~\cite{Angeby_estimating_2000}.  The signals are also used to describe the sounds emitted by bats for echo-location \cite{Peleg_DPT_1995}.  Of practical importance is the estimation of the coefficients $\beta = (\beta_0, \dots \beta_m)$ from $N$ consecutive samples $y_1, \dots. y_N$ where 
\[
y_n = s_n(\beta) + w_n
\]
and $\{w_n, n \in \ints\}$ is a sequence of noise variables.  Many estimators have been studied and implemented~\cite{Peleg_DPT_1995, Angeby_estimating_2000, Djuric_phase_unwrap_chirp_1990, Oshea_iterative_1996, Barbarossa_analysis_of_PPS_1997, Slocumb_polynomial_1994, Morelande_bayes_unwrapping_2008, Kitchen_polyphase_unwrapping_1994}.

Under the assumption that $\{w_n\}$ are additive white and Gaussian with variance $\sigma^2$, Peleg and Porat~\cite{Peleg1991_CRB_PPS_1991} have derived the Cram\'{e}r-Rao lower bound for unbiased estimators of the coefficients $\beta$.  Their approach requires inversion of a poorly conditioned Fisher information matrix.  To avoid this, they provide an approximation that is valid when the number of observations $N$ is sufficiently larger than the order of the polynomial phase signal $m$.  Here we remove the need for this approximation.  By converting the polynomial basis to one that is orthogonal we derive closed form formula for the inverse Fisher information matrix.  This leads to closed form formula for the Cram\'{e}r-Rao bounds.  %At the same time we tie together some existing literature on the Cram\'{e}r-Roa bound for polynomial phase signals~\cite{Peleg1991_CRB_PPS_1991,Ristic_crbpelegcomment_1998}.

The paper is organised as follows.  In Section~\ref{sec_crb} we derive the Fisher information matrix in a similar way to Peleg and Porat~\cite{Peleg1991_CRB_PPS_1991}.  The matrix is poorly conditioned and inverting it to obtain the Cram\'{e}r-Rao bound is difficult.  In Section~\ref{sec:changing-basis} we describe some general results regarding changing the basis in which a polynomial phase signal may be represented.  We describe the Fisher information matrix corresponding with a chosen basis, and we relate the matrices between two bases.  In Section~\ref{sec:discr-orth-polyn} we introduce the family of discrete orthogonal polynomials.  By transforming into this basis, the corresponding Fisher information matrix becomes diagonal, making matrix inversion simple.  This leads directly to closed form expressions for the Cram\'{e}r-Rao bounds.  To keep our derivation simple we initially assume that the amplitude $\alpha$ and the noise variance $\sigma^2$ are fixed constants, and not parameters to be estimated.  However, including the amplitude and noise variance is straightforward, and we do this in Section~\ref{sec:including-amplitude}.  Section~\ref{sec:simuations} presents the results of Monte-Carlo simulations that agree with our exact bounds.

@@

We make the assumption that the noise sequence $\{ w_n \}$ is independent and identically distributed complex Gaussian, each element having independent real and imaginary parts with zero mean and variance $\sigma^2$.  The probability density function of $w_n$ is 
\[
p(x) = \frac{1}{2 \pi \sigma^2} \exp\left( -\frac{\abs{x}^2}{2\sigma^2} \right)
\]
where $\abs{x}^2$ is the squared magnitude of the complex number $x$.  We wish to estimate $\beta$ from $y_1, \dots, y_N$.  The likelihood function is
\[
L(b) = \frac{1}{(2 \pi \sigma^2)^N} \exp\left( - \frac{1}{2\sigma^{2}} \sum_{n=1}^N \abs{y_n -  s_n(b) }^2 \right).
\]
%The maximum likelihood estimators of $\alpha, \beta_0, \dots, \beta_m$ are the maximisers of $L$.
% , that is
% \[
% (\hat{\alpha}, \hat{\beta}_0, \dots, \hat{\beta}_m) = \arg\max_{a \in \reals, (b_0, \dots, b_m) \in B} L(a, b_0, \dots, b_m) 
% \]  
% where $B \subset \reals^{m+1}$ describes a region of polynomial phase coefficients for which the estimation problem is identifiable~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase,McKilliam_LSU_polyest_part1_2012}.  We assume that $(\beta_0, \dots, \beta_m) \in B$.
For $i,k \in \{0, \dots, m\}$ define functions 
\[
f_{ik}(b) = f_{ki}(b) = -\expect \frac{ \partial^2 \log L(b) }{ \partial b_{i} \partial b_{k}},
\]  
where $\expect$ denotes expectation.  The Fisher information matrix $F$ is the $m+1$ by $m+1$ matrix with elements $F_{i k} = f_{ik}(\beta)$.  The elements are
\[
F_{ik} = F_{ki} = \frac{\alpha^2}{\sigma^2}\sum_{n=1}^{N} n^{i}n^{k}.
\]
It is a convenient property of polynomial phase estimation that the Fisher information matrix does not depend on the value of the polynomial phase coefficients being estimated.  That is, $F$ does not depend on $\beta$.

Let $\hat{\beta}$ be an unbiased estimator of $\beta$.  The Cram\'{e}r-Rao bound asserts:  If $C$ is the covariance matrix of $\hat{\beta}$, then $C - F^{-1}$ is positive semidefinite, where $F^{-1}$ is the inverse of $F$.  The diagonal elements of a positive semidefinite matrix are nonnegative so $C_{ii} \geq F^{-1}_{ii}$ and correspondingly, 
\[
C_{ii} = \var \hat{\beta}_{i} \geq F_{ii}^{-1}, \qquad i=0,\dots,m.
\]  
Be aware that $F_{ik}^{-1}$ is the $ik$th element of the inverse matrix $F^{-1}$.  It is not the reciprocal of $F_{ik}$.  Peleg and Porat~\cite{Peleg1991_CRB_PPS_1991} found that $F$ is poorly conditioned and for this reason $F^{-1}$ is difficult to compute.   We will circumvent this problem by applying a change of basis to the polynomial coefficients $\beta$.  This will lead to closed form formula for $F^{-1}$.

@@

The likelihood function for $\gamma$ is 
\[
L_\gamma(c) = L(cP) = L(b)
\] 
and we can compute the corresponding Fisher information matrix.  It is found to be the $m+1$ by $m+1$ matrix $H$ with elements
\[
H_{ik} = H_{ki} = \frac{\alpha^2}{\sigma^2}\sum_{n=1}^{N} p_{i}(n) p_{k}(n).
\]
The Cram\'{e}r-Rao bound for $\gamma$ is found by taking the inverse of $H$.  The next well know (even trivial) property relates the two Fisher information matrices $H$ and $F$.  We do not intend to claim any novelty by including a formal statement of this property.  The same property is, for example described in \cite[]{Kay1993_stat_sig_est_theory}.  We included a proof simply for completeness.

\begin{property}
Suppose we desire to estimate $\beta \in \reals^m$ and that $L(b)$ is the corresponding likelihood function and $F$ the corresponding $m \times m$ Fisher information matrix.  Let $P$ be an $m \times m$ invertible matrix and $\gamma \in \reals^m$ be such that $\beta = \gamma P$.  Then the likelihood function corresponding to $\gamma$ is $L_\gamma(c) = L(cP)$ and the corresponding Fisher information matrix is $H = P F P^\prime$. 
\end{property}
\begin{IEEEproof}
Put $b = cP$.  That $L_\gamma(c) = L(cP) = L(b)$ is plain.  The elements $F_{ik} = f_{ik}(\beta)$ where
\[
f_{ik}(b) = -\expect \frac{\partial \log L(b)}{\partial b_i \partial b_k}.
\]
The elements $H_{ik} = h_{ik}(\gamma)$ where
\[
h_{ik}(c) = -\expect \frac{\partial \log L_\gamma(c)}{\partial c_i \partial c_k} =  -\expect \frac{\partial \log L(cP)}{\partial c_i \partial c_k}.
\]
Since $\frac{\partial b_s}{\partial c_i} = P_{is}$ applying the chain rule yields
\begin{align*}
h_{ik}(c) &= -\expect \sum_{s=0}^{m-1} P_{is}\frac{\partial \log L(cP)}{\partial b_s \partial c_k} \\
&= -\expect \sum_{s=0}^{m-1}\sum_{t=0}^{m-1} P_{is} P_{kt}   \frac{\partial \log L(b)}{\partial b_s \partial b_t} \\
&= \sum_{s=0}^{m-1}\sum_{t=0}^{m-1} P_{is} P_{kt}  f_{st}(b).
\end{align*}
Putting $c = \gamma$ so that $b = \gamma P = \beta$ we have
\[
H_{ik} = \sum_{s=0}^{m-1}\sum_{t=0}^{m-1} P_{is} P_{kt}  F_{st},
\]
or equivalently $H = P F P^\prime$.
\end{IEEEproof}
